---
title: "tutorial 1"
output: html_document
date: "2026-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("network")
#install.packages("igraph")
#install.packages("reshape2")
#install.packages("tidyverse")
library(network)
library(igraph)
library(reshape2)
library(tidyverse)
```
```{r}
url1 <- "https://github.com/rensec/sasr08/raw/main/g_adj_matrix_simple.csv"
g_matrix <- read.csv(file = url1, header = FALSE)
g_matrix
```
```{r}
class(g_matrix)
```
```{r}
g_matrix <- as.matrix(g_matrix)
g_matrix
```
```{r}
rownames(g_matrix) <- 1:nrow(g_matrix)
colnames(g_matrix) <- 1:ncol(g_matrix)
g_matrix
```
```{r}
g_nodes_age <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_nodes_age.csv")
g_nodes_age
```
```{r}
g <- graph_from_adjacency_matrix(g_matrix)
class(g)
```
```{r}
g
```
This output shows you that R sees the matrix g as a network; it shows you the edges.

```{r}
plot(g)
```

```{r}
g <- set_vertex_attr(g, 
                     name = "age",
                     value = g_nodes_age$age)
```
```{r}
plot(g)
```
QUESTION: I dont really understand what we did here?

```{r}
g_elist <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_elist.csv")
g_elist
```
```{r}
g_from_elist <- graph_from_data_frame(g_elist) 
plot(g_from_elist)
```
```{r}
g_from_elist <- graph_from_data_frame(g_elist, vertices = g_nodes_age) 
plot(g_from_elist)
```

1.3; adjacency list -> to igraph (network) object
```{r}
g_adj_list <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_adj_list.csv")
g_adj_list
```
```{r}
g_elist_from_alist <- g_adj_list
library(igraph)
g_elist_from_alist <-  select(g_elist_from_alist, id, friend1:friend2) # keep only the network variables (and id)
g_elist_from_alist <-  melt(g_elist_from_alist, id.vars = "id") # from the reshape2 package
g_elist_from_alist
```
```{r}
g_elist_from_alist <- g_elist_from_alist%>% 
  filter(!is.na(value)) %>% # drop the missings
  rename(from ="id", to = "value", sourcevar= "variable") %>% #just nice for interpretation
  relocate(to, .after=from) #move around the columns
g_elist_from_alist
```
With this you make the adjacency matrix more interpretable.
--> sourcevar is an edge attribute indicating where in the adjacency list columns the alter was mentioned. Strictly speaking you don’t need the latter for a simple edge list, but it’s useful to keep in case you want to reverse the procedure.
```{r}
g_from_alist <- graph_from_data_frame(g_elist_from_alist) #This is a proper igraph graph object
plot(g_from_alist)
```
we lost the isolate, so we need to include.
```{r}
nodelist <- select(g_adj_list,id,age)
g_from_alist <- graph_from_data_frame(g_elist_from_alist, vertices = nodelist) #This is a proper igraph graph object
plot(g_from_alist)
```

reverse from igraph object -> adjacency matrix
```{r}
g_adj_matrix <- as_adjacency_matrix(g, sparse = FALSE)
g_adj_matrix
```
```{r}
all(g_adj_matrix == g_matrix)
```
You do this to check whether the reversed matrix is indeed the same as the one we got from the file.

reverse igraph object -> elist
```{r}
g_edgelist <- igraph::as_data_frame(g_from_elist, what = "edges") 
g_edgelist 
```
Note that now, once more, we have lost our isolated node 5, as it is not included in the edge list!
```{r}
igraph::as_data_frame(g_from_elist, what = "vertices") 
```

from igraph object -> adjacency list
```{r}
edgelist_2 <- igraph::as_data_frame(g_from_alist, what = "edges")
nodelist_2 <- igraph::as_data_frame(g_from_alist, what = "vertices")
d <- edgelist_2 %>%
  rename(id = "from") %>% 
  pivot_wider( id_cols = id, names_from = sourcevar, values_from = to ) %>% 
  merge(nodelist_2, by.x = "id", by.y = "name", all.y = TRUE ) %>% 
  relocate(age, .after = id) %>% 
  type_convert() #as_data_frame returns characters; this transforms it back to numeric
```
```{r}
d
```


```{r}
g_np <- as.network(g_elist_from_alist, matrix.type = "edgelist", vertices = nodelist)
g_np
```
```{r}
plot(g_np)
```

The choice between 'network' and 'igraph' is just depending on your research question, but igraph has my preference up until this point.

```{r}
g_mod <- delete_vertices(g,2)
plot(g_mod)
```
```{r}
g_mod <- delete_vertices(g,which(V(g)$age == 21))
plot(g_mod)
```
here you remove all the nodes with the age 21.
```{r}
g_mod <- delete_vertices(g, which(is.na(V(g)$age)))
plot(g_mod)
```
```{r}
V(g)$gender <- c("male", "female", "female", "other", "male") 
g
```
This is an alternative way to add attibrutes to nodes.

Other fun things:
```{r}
g_rev <- reverse_edges(g)
plot(g_rev)
```
```{r}
g_union <- igraph::union(g, g_rev)
```
```{r}
plot(g_union)
```
This is basically an undirected network. 
```{r}
g_undir <- as_undirected(g)
plot(g_undir)
```

