---
title: "Tutorial 4"
output: html_document
date: "2026-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(klippy)
library(remotes)
library(igraph)
library(haven) # to read Stata files
library(reshape2)
library(tidyverse)
library(knitr)
library(network)

```
```{r}
install.packages("spdep")
```
```{r}
library(spdep)
data(columbus)
mydata <- columbus
```
```{r}
attach(mydata)
```

Question: I dont get why you need to do this... what do you mean use dollar sign? or is this for further in the analysis?

```{r}
install.packages("spatialreg")
library(spatialreg)
```
```{r}
neighbors <- col.gal.nb
```
```{r}
class(neighbors)
```
This means that the dataframe is a neighborhood object. For spatial matrix you need a ‘weighted list - listw’ object. 
1. first convert the dataframe to nb; class(dataframe) <- “nb”
2. then do the following:
```{r}
listw <- nb2listw(neighbors, style = "W")
summary(listw)
```
the average number of links: 4.69. So every region has an average of 5 neighbors. 

```{r}
summary(CRIME)
summary(INC)
summary(HOVAL)
```
```{r}
plot(neighbors, coords)
```
R needs the coordinates of the neighborhoods to know where to place them, this is what chat says: 
"The coords argument provides the actual positions (x, y) of each observation so that:

Each node can be placed on the map
Edges (connections) can be drawn between the correct locations

Without coordinates, R would know the connections but wouldn’t know where to draw them."

```{r}
olsreg <- lm(CRIME ~ INC + HOVAL)
summary(olsreg)
```
The coefficients show a negative association between income and crime rate in a neighborhood. Only we did not take spacial dependence into account, so we need to check whether this has to be done:

Moran's I test:
```{r}
moran.test(CRIME, listw)
```
Is significant.
Now we need to figure out which type of spatial models we should use. For this purpose, we use Lagrange multiplier tests for spatial lag and spatial error dependencies.

Lagrange multiplier test:
```{r}
lm.LMtests(olsreg, listw, test=c("LMlag", "LMerr"))
```
The null-hypotheses of no spatial error dependence and no spatial lag dependence can both be rejected; so there is spatial dependence.
--> remember that these hypotheses are not compared to each other, but only to linear regression.
--> use both models (lag and error) and see whether results differ.

The spatial lag model (spillover effects):
```{r}
slm <- lagsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = "FALSE")
summary(slm)
```
Do not interpret the b, for there is an autoregressive term in the formula; the y is predicting y itself...
--> Rho: spatial parameter is significant, showing us that there is spatial lag dependency. 

Direct and indirect impacts:
```{r}
im<-impacts(slm, listw = listw, zstats=TRUE, R=200)
sums<-summary(im,  zstats=T)
data.frame(sums$res)
```
specify zstats for asking the p-values.
Question: what is res?? If it is residual, how does this give me the direct and indirect impacts?
```{r}
data.frame(sums$pzmat)
```
The direct and indirect effect are both (almost) significant, however, the total effect is mostly driven by the direct effect (also including the feedback loop). 

Now the spatial error model (question: exogenous effects?)
```{r}
sem <- errorsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = FALSE)
summary(sem)
```
Lamba (the spatial parameter) is again positive and significant and the regression coefficients are again negative and significant.

Now the spatial autoregressive combined model:
```{r}
sac <- sacsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = FALSE)
summary(sac)
```
Rho is not significant... because small dataset.

spatial durbin model: exogenous effects (characteristics of the neighbors, not the outcome of the neighbor)
```{r}
sdm <- lagsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = TRUE)
summary(sdm)
```
the lag's of the X's are not significant, but rho is significant. Although you cant interpret the lag coefficients, you can say that the characteristics of the neighbors do influence the outcome (level of crime) of the other neighborhood.

Then the spatial durbin error model:
```{r}
sdem <- errorsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = TRUE)
summary(sdem)
```
General nested model:
```{r}
gnm <- sacsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = TRUE)
summary(gnm)
```
Again, you have too small a dataset to take conclusions based on the lag coefficients.

Assignment 4.2
```{r}
PupilsWaveV1 <- read_dta(file = "PupilsWaveV (1).dta")
```
Q2 Estimate a linear regression model of doing homework (acthomew) on sex (sex) while controlling for age (age)
--> reverse the coding of acthomew and sex

```{r}
class(PupilsWaveV1)
```
```{r}
attach(PupilsWaveV)
```
This enables you to just ask for the variables, instead of always using brackets.
```{r}
table(acthomew)
table(sex)
```
```{r}
str(sex)
```

So acthomew is measured on a scale from 1-5, with missing 9
And sex on 1 female and 2 male.
Now recode these.
```{r}
PupilsWaveV1$sex[PupilsWaveV1$sex == 9] <- NA
acthomew_r <- 6 - acthomew
female <- ifelse(sex == 2, 0,
                 ifelse(sex == 1, 1, NA))
```

```{r}
table(acthomew_r)
table(female)
```
```{r}
olsreg <- lm(acthomew_r ~ female + age)
```
```{r}
summary(olsreg)
```
Q3 Estimate a multilevel linear model where you take class characteristics into account.

```{r}
#install.packages("lme4")
library(lme4)
```

```{r}
model <- lmer(acthomew_r ~ female + age + (1 | schoolnr), data = PupilsWaveV1)
```
```{r}
summary(model)
```

```{r}
class(PupilsWaveV1)
```

Q4. In addition to gender differences, we are now also interested in whether peers influence one’s homework behavior. From now on consider only Class 12b. Construct an adjacency matrix of best friends.
```{r}
class12b <- PupilsWaveV1 %>% 
  filter(schoolnr == "12b")  # keep only one class
```
data is in the format of adjacency list, now transform to matrix (i transformed it to edge list; object) of best friends. 

```{r}
knecht_4b <- select(class12b, namenr, friend1:friend12) # keep only the network and variables (and id)

```
```{r}
library(igraph)
library(reshape2)
library(tidyverse)
g_elist_from_alist <-  melt(knecht_4b, id.vars = "namenr") # from the reshape2 package
g_elist_from_alist
```
```{r}
g_elist_from_alist <- g_elist_from_alist%>% 
  filter(!is.na(value)) %>% # drop the missings
  rename(from ="namenr", to = "value", sourcevar= "variable") %>% #just nice for interpretation
  relocate(to, .after=from) #move around the columns
g_elist_from_alist
```

Now i can also transform it to an adjacency matrix.
```{r}
#g_adjmatrix_from_elist <- graph_from_data_frame(g_elist_from_alist, directed = FALSE)
#mat <- as_adjacency_matrix(g_adjmatrix_from_elist)

```
```{r}
g_from_alist <- graph_from_data_frame(g_elist_from_alist) #This is a proper igraph graph object
plot(g_from_alist)
```

Q5. Construct a row-standardized spatial weight matrix as a weights list object. Set the option ‘zero.policy’ to TRUE because you have isolates.
First change the dataframe to a weight list object.

```{r}
class(mat)
```
```{r}
library(spdep)
listw <- mat2listw(mat, style = "W", zero.policy = TRUE)
```
So, to make a weighted list you need the adjacency matrix of best friends. Which is why i first made this (Q4). 
```{r}
summary(listw)
```
Approximately, a pupil has 5 ties.

Q6. Conduct and interpret the Moran’s I test and the Lagrange multiplier tests. Moran’s I expects a weights list object.

```{r}
moran.test(acthomew_r, listw)
```
Error in na.fail.default(x) : missing values in object

```{r}
library(spatialreg)
```
```{r}
moran.test(acthomew_r, listw, na.action = na.omit)
```
Error in subset.listw(listw, subset, zero.policy = zero.policy) :
Not yet able to subset general weights lists
solution: you have to report the missings in homework behavior and then remake the weighted list? -> takes too much time for me to do now.
```{r}
listw_clean <- spdep::subset(listw, subset = keep, zero.policy = TRUE)
```
```{r}
#keep <- complete.cases(acthomew_r)

#x_clean <- acthomew_r[keep]
```













