---
title: "tutorial 2"
output: html_document
date: "2026-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(igraph)
z <- make_graph("Zachary")
z
```
```{r}
plot(z)
```

```{r}
is_directed(z)
```
This is to check whether the network is undirected. And it is not...

First you have to ask for the so-called five number summary.
1) size
```{r}
vcount(z)
ecount(z)
```
2) density: the ratio of existing edges and the maximum possible number of edges that could exist, given the number of vertices
```{r}
edge_density(z)
```
```{r}
z$density <- edge_density(z)
z
```
‘Density’ is now listed as one of the attributes of the object, where ‘g/n’ indicates that it is a graph-level attribute. You can also do this for the other ones.

3) components: the number of unconnected parts of the network (which may be parts consisting of one node, that is, isolates).
```{r}
count_components(z)
```
4) distance; the longest shortest path; Among all the shortest paths, take the longest one.
```{r}
diameter(z)
```
```{r}
mean_distance(z)
```
5) clustering coefficient:  the extent to which triangles tend to be closed in the network
```{r}
transitivity(z, type="average")
```
for each vertex, the proportion of the potential ties between the vertex’ neighbors that actually exist (in the ego networks literature, this is referred to as local density), and averages this over all vertices

```{r}
V(z)$degree <- degree(z)
plot(z, vertex.size = V(z)$degree+8, margin= -0.1)
```

For the other centrality measures: see tutorial 
- betweenness centrality
- closeness centrality




