---
title: "Tutorial 5"
output: html_document
date: "2026-05-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(network)
library(intergraph)
library(sna)
library(igraph)
install.packages("ergm")
library(ergm)
library(tidyverse) 
library(haven)
```
```{r}
load(url("https://github.com/rensec/sasr08/raw/main/classnet.rdata"))
summary(classnet)
```
Unfortunately, network does not handle factor variables, so we first need to convert ‘sex’ to a numeric variable.

```{r}
V(classnet)$sex <-as.numeric(V(classnet)$sex)
g <- asNetwork(classnet)
g
```
```{r}
gplot(g,  vertex.col=g %v% "sex"*2)
```
Looks like we have some gender homophily going on here, as well as a lot of reciprocity! With ERGM, we can test this statistically.

Testing the hypothesis: ties are likely to be reciprocated.
```{r}
M1.g<-ergm(g ~ edges +  mutual )
```
question: where do you find this list of tendencies i could use? because i wouldnt know myself that i could just use mutual as an independent var.

```{r}
summary(M1.g)
```
Edges is significant: tells u something about the random probability of forming ties.
Mutual is significant: when there already is a tie (directed), the probability of a reciprocated tie (directed back) is higher. 

When you add the nodematch effect (homophily of genders) and the nodefactor effect (difference between genders in their ties):

```{r}
M2.g<-ergm(g ~ edges + mutual +  nodefactor("sex", levels = 2) + nodematch("sex" ))

summary(M2.g)
```
nodematch is significant, so genders are more likely to have ties with the same gender
nodefactor is not significant, so boys do not have more ties than the other.

```{r}
M2.g.gof <- gof(M2.g)
M2.g.gof
```
If the observed value (obs) falls within the range (min-max) then the model fits your data well.
the p-value is a Monte Carlo test: “How extreme is the observed value compared to what the model typically produces?”
--> close to one, the observed value is typical for this model.
the model statistics (what the model measures) looks good.
the features that were not included in the model, are not typical under this model: 
chat --> So you look at:

Degree distribution
Distances
Clustering patterns (ESP, triangles, etc.)

If you see:

Observed line far outside simulated envelopes, or
Very low p-values

👉 Then the model is missing structure
```{r}
plot(M2.g.gof)
```
for directed networks, it is common to include at least some parameters related to degree, reciprocity, and something related to triadic closure. 
--> the fact that the goodness of fit is quite low for the unmeasured features, says that you should include these in the model. 


